A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-09-13 submission/manuscript.rmd
update abstract

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics
(OHDSI) initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
library of open-source tools with a framework for risk-based assessment of
treatment effect heterogeneity. The proposed framework consists of five
steps: 1) definition of the problem, i.e. the population, the treatment, the
comparator and the outcome(s) of interest; 2) identification of relevant
databases; 3) development of a prediction model for the outcome(s) of
interest; 4) estimation of propensity scores within strata of predicted risk and
estimation of relative and absolute treatment effect within strata of predicted
risk; 5) evaluation and presentation of results. We demonstrate our framework by
evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE)
inhibitors versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.05% to -0.02% in the lowest risk quarter to 
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. The proposed framework may be useful for the evaluation of
heterogeneity of treatment effect on observational data that are mapped to the
OMOP Common Data Model. The proof of concept study demonstrates its feasibility
in large observational data. Further insights may arise by application to safety
and effectiveness questions across the global data network.

>>>>

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics
(OHDSI) initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
library of open-source tools with a framework for risk-based assessment of
treatment effect heterogeneity. The proposed framework consists of five
steps: 1) definition of the problem, i.e. the population, the treatment, the
comparator and the outcome(s) of interest; 2) identification of relevant
databases; 3) development of a prediction model for the outcome(s) of
interest; 4) estimation of relative and absolute treatment effect within strata
of predicted risk, after adjusting for observed confounding; 5) presentation of
the results. We demonstrate our framework by evaluating heterogeneity of the
effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on
3 efficacy and 6 safety outcomes across three observational databases. Patients
at low risk of acute myocardial infarction (MI) received negligible absolute
benefits for all 3 efficacy outcomes, though they were more pronounced in the
highest risk quarter, especially for hospitalization with heart failure. The
substantial risk increase of cough and angioedema with ACE inhibitors across all
risk strata suggests that beta blockers provide a viable alternative to patients
at low risk of acute MI. The proof of concept study demonstrates its feasibility
in large observational data. Further insights may arise by application to safety
and effectiveness questions across the global data network.
